---
title: "Code Companion: Mastering R for Data Science 2026"
author: "Vadim Tyuryaev"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    theme: paper
    highlight: tango
    code_folding: show
    code_download: true
    df_print: paged
    css: styles.css
header-includes:
  - |
    <meta name="description" content="Mastering R for Data Science: Essential Tips, Hidden Tricks, and Pro-Level Coding Workflows 2026.">
    <meta name="keywords" content="R Programming, High-Performance Computing, Parallel Computing, GPU Acceleration, Neural Networks, NVIDIA, SVD, Reticulate, TensorFlow, Keras, Data Science.">
    <meta name="author" content="Vadim Tyuryaev">
    <meta name="robots" content="index, follow">
    <meta property="og:title" content="R-Cyclopedia: The Ultimate R Guide">
    <meta property="og:description" content="The Ultimate R Guide: Vectorization, Python Integration, Parallel Computing, and GPU-based Deep Learning with TensorFlow.">
    <meta property="og:type" content="article">
    <meta property="og:url" content="https://vadimtyuryaev.github.io/R-Cyclopedia">
    <meta property="og:image" content="https://vadimtyuryaev.github.io/R-Cyclopedia/preview-image.png">
    <meta property="og:site_name" content="Vadim Tyuryaev’s Statistical Blog">
---

```{r setup, include=FALSE}
# Global chunk options: evaluate code to show outputs, suppress messages/warnings
knitr::opts_chunk$set(
  eval      = TRUE,   
  echo      = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

---

# Basics

## Assign and Retrieve Values

Utilize `assign()` and `get()` in loops when dealing with a dynamic number of 
variables.

```{r basics-assign}

# assign the value of 2 to a newly created variable called var.1
assign(paste("var", 1, sep="."),2)

# retrive the value of the var.1 variable
get("var.1")

```
```{r basics-assign-loop}

set.seed(as.numeric(as.Date("1777-04-30")))

# Create a data frame directly
my_data = data.frame(
  id = paste("rval", 1:5, sep = "_"),
  value = rnorm(5)
)

print(my_data)

```

## Environment Scoping

Create custom environments to control where R looks for variables, leveraging
parent-child links.


```{r basics-env-parent-child}

defaults      = new.env()
defaults$tax  = 0.15

order         = new.env(parent = defaults)
order$price   = 200

# 'tax' is not in 'order', but is found via its parent environment
order$total   = order$price * (1 + get("tax", envir = order))
order$total

```
```{r env-parent-child}

# Create a parent environment
parent.env = new.env()

# Add some variables to the parent environment
parent.env$a = 10
parent.env$b = 5

# A simple calculation in the parent environment
parent.env$sum_parent = parent.env$a + parent.env$b   # 15

# Check what's in the global environment (you should NOT see a, b, or sum_parent)
ls()                # default is ls(globalenv())

# Check what's inside parent.env
ls(parent.env)
# [1] "a" "b" "sum_parent"

# Try to access 'a' from the global environment (this should give an error)
try(a, silent = FALSE)
# Error: object 'a' not found

# Correct way: explicitly get 'a' from parent.env
get("a", envir = parent.env)        # 10

```

## Evaluate Expressions

Dynamically construct and evaluate expressions from strings, or extract data from model calls.

```{r basics-eval}

# Define an expression
expr = expression(sqrt(81))

# Evaluate the expression
result = eval(expr)

# Print the result
print(result)

```

```{r basics-eval-parse}

# Dynamically construct an expression
x = 12
operation = "x ** 2"
expr = parse(text = operation)

# Evaluate the dynamically constructed expression
result = eval(expr)

# Print the result (returns 144)
print(result)

```

```{r basics-eval-lm}

set.seed(42)

lm.data  = data.frame(x = 1:10, y = 2.5 * (1:10) + rnorm(10))
lm.model = lm(y ~ x, data = lm.data)

# Extract the original data frame via the stored call
head(eval(lm.model$call$data), 3)

```

## For Loops and Preallocation

```{r basics-for-loops}

# basic usage
for (i in 1:10){
  print(i)
}

```

```{r}

library(tictoc)

set.seed(1905)

n_sims = 100000

sim_means_grow = numeric(0)               # Empty vector

tic("Without pre-allocation")
for (i in seq_len(n_sims)) {
  sim_means_grow = c(sim_means_grow, mean(rnorm(50)))
}
toc()



sim_means_prealloc = numeric(n_sims)      # Pre-allocated vector

tic("With pre-allocation")
for (i in seq_len(n_sims)) {
  sim_means_prealloc[i] = mean(rnorm(50))
}

toc()

```

> *Always preallocate storage before calculating for-loops to drastically improve efficiency*.

```{r basics-for-bench}

set.seed(1925)

library(microbenchmark)

n = 1e4 

bench = microbenchmark(
  prealloc = { 
    v = numeric(n)
    for (i in seq_len(n)) v[i] = i     # EFFICIENT 
  },
  grow = { 
    v = numeric(0)
    for (i in seq_len(n)) v = c(v, i)  # NOT EFFICIENT
  },
  times = 10
)

# 1. Save the filtered summary to a data frame
bench_summary = summary(bench)[, c("expr", "median", "min", "max")]

# 2. Round the numeric columns (columns 2, 3, and 4) to 0 decimal places
bench_summary[, 2:4] = round(bench_summary[, 2:4], digits = 0)

# 3. Print the clean results
print(bench_summary)

```

## Memory & Path Management

```{r basics-memory, eval=FALSE}

rm(list = ls())      # Clear current environment
gc()                 # Garbage collection
.rs.restartR()       # Restart R session (RStudio only)

getwd()              # Get current working directory
setwd("..")          # Go one level up in the directory tree
setwd("~")           # Go directly to the user's home directory

```

## RDS format

```{r matrix-rds}

# Create a sample 3x3 matrix
sample_matrix = matrix(1:9, nrow = 3, ncol = 3)
print(sample_matrix)

# Save the matrix as an .rds file in the current directory
saveRDS(sample_matrix, file = "./sample_matrix.rds")

# Read the file back into a new variable to verify
restored_matrix = readRDS("./sample_matrix.rds")

# Check if the restored object is identical to the original
identical(sample_matrix, restored_matrix)

```

## Advanced Indexing

Use `which()` with `arr.ind = TRUE` to get row and column indices in data frames.

```{r basics-which}
x = c(3, 7, 2, 6, 7, 9, 1)

which(x == 7)
which(x > 5)
x[which(x > 5)]

```

```{r basics-which-min-max}

# indexes of maximum and minimum
which.max(x) # [1] 6
which.min(x) # [1] 7

# Get the actual values 
x[which.max(x)] # [1] 9
x[which.min(x)] # [1] 1

```
```{r basics-which-arrind}

df = data.frame(a = c(1, 3, 5), 
                 b = c(5, 3, 1)
                 )

print(df)

which(df > 3, arr.ind = TRUE)

```

---

# Functions

## Returning Multiple Objects

Group multiple outputs (data, models, plots) into a single nested list.

```{r functions-multi-return}

# Define sample function with input parameters
my_function = function(x, y){
  
  # Perform some operations
  result1 = x^2
  result2 = y^3
  
  # Perform additional operations
  result3 = x^2+y^2
  
  # Combine results into a list
  result_list      = list(
    x_squared      = result1,
    y_cubed        = result2,
    sum_of_squares = result3
  )
  
  # Return the list of results
  return(result_list)
}

# Example usage of the function
output = my_function(3, 5)

# Access the third result (sum of 3^2 + 5^2)
output$sum_of_squares

```

## Optional Arguments with Ellipsis (`...`)

Pass varied arguments to internal functions using `...` .

```{r functions-ellipsis}

# Define the main plotting function
plot_modern = function(x, y, ...) {
  
  # Internal function to create a stylized plot
  internal_plot = function(x, y, ...) {
    
    # 1. Set a soft background color for the plot area
    par(bg = "#f4f7f6", mar = c(5, 5, 4, 2))
    
    # 2. Initialize an empty plot window (type = "n") to lay down the grid first
    plot(x, y, type = "n", bty = "n", ...)
    
    # 3. Add a clean white grid 
    grid(nx = NULL, ny = NULL, col = "white", lty = 1, lwd = 1.5)
    
    # 4. Draw a smooth trendline underneath the points
    lines(spline(x, y), col = "#b0bec5", lwd = 2)
    
    # 5. Plot the actual points, passing the ellipsis (...) for custom styling
    points(x, y, ...)
  }
  
  # Execute the internal function
  internal_plot(x, y, ...)
}


# 1. Generate sample data
set.seed(42)
x = 1:15
y = x^2 + rnorm(15, mean = 0, sd = 15)

# 2. Draw the standard base R plot (Left side)
plot(x, y, 
     main = "Standard Base R Plot",
     xlab = "X", ylab = "Y")

# 3. Draw our custom modern plot (Right side)
plot_modern(x, y, 
            main = "Quadratic Growth Trajectory", 
            xlab = "Time (X-axis)", 
            ylab = "Value (Y-axis)", 
            col  = "#ffffff",   # Point border color
            bg   = "#ff6b6b",   # Point fill color (works with pch = 21)
            pch  = 21,          # Filled circle 
            cex  = 2,           # Point size
            lwd  = 1.5,         # Border thickness
            family = "sans")    # Modern font

```

## Multiple optional arguments with `do.call()`.

```{r functions-do-call}

# Define the main wrapper function
process_and_plot = function(data, plot_args = list(), ...) {
  
  # Internal: Calculate statistics using arguments from the ellipsis (...)
  calculate_summary = function(data, ...) {
    summary(data, ...)
  }
  
  # Internal: Create a histogram with modern default styling
  create_histogram = function(data, main = "Histogram", xlab = "Values", 
                               col = "#90CAF9", border = "white", breaks = 15) {
    hist(data, main = main, xlab = xlab, col = col, border = border, breaks = breaks)
  }
  
  # 1. Execute summary via do.call, passing along the ellipsis arguments
  statistics = do.call(calculate_summary, list(data = data, ...))
  
  # 2. Inject data into plot_args and execute the histogram
  plot_args$data = data
  do.call(create_histogram, plot_args)
  
  # Implicitly return the statistics object
  statistics
}

# Example usage
set.seed(1812)
sample_data = rnorm(100)

# Define custom arguments to override the histogram defaults
custom_plot_args = list(
  main   = "Customized Distribution", 
  col    = "#5C6BC0", 
  border = "white", 
  breaks = 25
)

# Call the function: 
# 'digits' and 'quantile.type' flow into the ellipsis (...) for the summary
# 'custom_plot_args' flows into the histogram
example_stats = process_and_plot(
  sample_data, 
  plot_args     = custom_plot_args,
  digits        = 2, 
  quantile.type = 5
)

# Print the resulting summary table
print(example_stats)

```


---

# Data Structures and Matrix Math

## Modifying Data Frames

Recreate the structure of a data frame and populate it cleanly.

```{r df-manipulation}

# Create a sample data frame with some initial data
sample_df = data.frame(
  ID     = c(1, 2, 3),
  Name   = c("Alice", "Bruce", "Candice"),
  Age    = c(25, 30, 35),
  Salary = c(50000, 55000, 60000)
)

# Print the sample data frame
sample_df

```

```{r df-null}

empty_df = sample_df[NULL, ]
empty_df

```

```{r df-direct-assignment}

# Fill the empty data frame by direct assignment
empty_df[1, ] = list(ID = 4, Name = "David", Age = 28, Salary = 55000)
empty_df[2, ] = list(ID = 5, Name = "Eva", Age = 32, Salary = 62000)

empty_df

```

## Matrix Multiplication & Logistic Regression

```{r maxtrix-multiplication-basics}

# set seed
set.seed(1918)

# Define matrix A(2x3)
A = matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3)
print(A)

# Define matrix B(3x2)
B = matrix(c(7, 8, 9, 10, 11, 12), nrow = 3, ncol = 2)
print(B)

# Perform regular matrix multiplication(2x2)
C = A %*% B
print(C)

# Define matrix D(2x2) and fill it by rows
D = matrix(c(2, 4, 6, 8), nrow = 2, ncol = 2, byrow=T)
print(D)

# Perform element-wise multiplication(2x2)
F = C * D
print(F)

```

```{r matrix-nn-train}

simple_logistic_NN=function(X,y, W0,b0, num_epochs,learning_rate, verbose=F){
  
# X  -> n × num_features
# y  -> n × 1
# W0 -> num_features × 1
# b0 -> 1 × 1

  # define the sigmoid function
  sigmoid = function(z) {
   1 / (1 + exp(-z))
  }

  # obtain the number of data points and features
  n=dim(X)[1]
  num_features=dim(X)[2]

  # merge weights and biases and add a column of 1s into X
  W_matrix=matrix(rbind(W0,b0), nrow=num_features+1,ncol=1)
  X_matrix=matrix(cbind(X, rep(1,n)),nrow=n,ncol=num_features+1)
  
  # preallocate a data frame to save losses and a matrix to save gradients
  col_names = c("Epoch", "Loss") 
  df_loss = data.frame(matrix(NA, nrow = num_epochs, ncol = length(col_names))) 
  colnames(df_loss) = col_names
  Gradients = matrix(NA,nrow= num_features+1, ncol=num_epochs)
  
# Training loop
for (epoch in 1:num_epochs)
  {
  
  # compute the forward pass
  z = X_matrix%*% W_matrix 
  y_hat = sigmoid(z)
  
  # calculate loss using binary cross-entropy
  loss = -mean(y * log(y_hat) + (1 - y) * log(1 - y_hat))
  
  # save epoch and loss
  df_loss[epoch,]=c(epoch, loss)
  
  # compute gradients
  dW = t(X_matrix) %*% (y_hat - y) / n
  
  # save gradient
  Gradients[,epoch]=dW
  
  # update weights and bias
  W_matrix = W_matrix - learning_rate * dW
  
  if(verbose!=FALSE){
  # Print loss every verbose epochs
    if (epoch %% verbose == 0) {
      cat("Epoch:", epoch, "Loss:", loss,"Weights:", W_matrix[1:num_features,],
          "Bias:", W_matrix[num_features+1,], "\n")
                               }
                 }
   }
  
  l=list("W_est"     = W_matrix[1:num_features,], 
         "b_est"     = W_matrix[num_features+1,],
         "Loss_DF"   = df_loss, 
         "Gradients" = Gradients)
  return(l)
}

```

```{r matrix-nn-example}

set.seed(1986)

# Generate a synthetic dataset
n_examp_1 = 1000
X_examp_1 = matrix(rnorm(n_examp_1 * 2), ncol = 2)
y_examp_1 = as.numeric(X_examp_1[,1] + X_examp_1[,2] > 0)

# Initialize weights and bias
W0_examp_1 = matrix(runif(2), nrow = 2)
b0_examp_1 = runif(1)

logistic_NN_training_examp_1=simple_logistic_NN(X=X_examp_1, y=y_examp_1, 
                                                W0=W0_examp_1,b0=b0_examp_1, 
                                                num_epochs=1000,
                                                learning_rate=0.01, 
                                                verbose=100)

# Print final weights and bias
cat("Estimated Weights:", logistic_NN_training_examp_1$W_est, "\n")
cat("Estimated Bias:", logistic_NN_training_examp_1$b_est, "\n")

```

## Singular Value Decomposition (SVD) from Scratch

```{r matrix-svd}

svd_custom = function(A) {
  # Check if A is a matrix
  if (!is.matrix(A)) stop("Input must be a matrix")
  
  # Step 1: Compute A^T * A
  AtA = t(A) %*% A
  
  # Step 2: Eigen decomposition of A^T * A
  eigen_AtA = eigen(AtA)
  
  # Eigenvalues and eigenvectors
  eigenvalues = eigen_AtA$values
  V = eigen_AtA$vectors
  
  # Step 3: Compute singular values
  singular_values = sqrt(eigenvalues)
  
  # Step 4: Compute U matrix
  U = matrix(0, nrow = nrow(A), ncol = nrow(A))
  for (i in 1:nrow(A)) {
    if (singular_values[i] != 0) {
      U[, i] = (A %*% V[, i]) / singular_values[i]
    }
  }
  
  # Step 5: Construct Sigma matrix
  Sigma = diag(singular_values)
  
  # Return the SVD components
  list(U = U, Sigma = Sigma, V = V)
}

```


```{r matrix-svd-example}

set.seed(1903)

A = matrix(rnorm(9), nrow = 3, ncol = 3)
svd_result =svd_custom(A)

# UVT^t == A (evaluates to TRUE)
all.equal(svd_result$U%*%svd_result$Sigma%*%t(svd_result$V), A)

```

## Pre-allocate lists

```{r preallocate-lists}

# Preallocate the empty list with 3 elements
n    = 3
list = vector("list", length = n)

# Fill in the first element with another list
list[[1]] = list(a = 5, b = 8, c = 13)

# Fill in the second element with a data frame
list[[2]] = data.frame(Name = c("Zu", "Georges"), Age = c(3, 14))

# Fill in the third element with a vector
list[[3]] = c(21, 34, 55, 89, 144)

list[[1]]$a
list[[2]]$Age
list[[3]]

```

## Python-Like Dictionaries in R

Use environments for reference semantics and O(1) lookup.

```{r dict-env}

# First way using an environment 
dict_env   = new.env()

# Assign key-value pairs to the environment
dict_env$a = 2
dict_env$b = 7
dict_env$c = 1

# Add a new key-value pair
dict_env$d = 8

# Modify an existing value
dict_env$a = 2

# List all objects in the environment
ls(dict_env)

# Get a value by key
get("c", envir = dict_env)

```
```{r dict-lists}

# generates a set of 100 colors using the rainbow function.
cols = rainbow(100)

# create a sequence of the first 100 odd numbers starting from 1.
numb = seq(1,200,2)

# preallocate three lists to store the colors, the numbers, and the indices
col_vals = vector("list",100)
num_vals = vector("list",100)
ind_vals = vector("list",100)

# iterate
for(c in 1:length(numb)){
  col_vals[[c]] = cols[c]
  num_vals[[c]] = numb[c]
  ind_vals[[c]] = c
}

# retrieve the color associated with the number 157 ("#AD00FF")
col_vals[[which(num_vals==157)]]

# find the number that corresponds to the color "#AD00FF" (157)
num_vals[[which(col_vals=="#AD00FF")]]

#check if the color and the number are associated with the same index (TRUE)
ind_vals[[which(col_vals=="#AD00FF")]]==ind_vals[[which(num_vals==157)]]

```

## Vectors

```{r vectors-types, eval=FALSE}

nums   = c(1, 2, 3)               # numeric
chars  = c("a", "b", "c")         # character
logic  = c(TRUE, FALSE, TRUE)     # logical
seq_10 = 1:10                     # integer sequence
rep_pi = rep(pi, 3)               # repeat a value

```

```{r vectors-examples}

x = 1:5
y = seq(10, 50, 10)

x + y        # 11 22 33 44 55
x^2 + log(y) # combined numeric ops

vals = c(4, 7, 11, 3, 9)

# Keep elements > 5
vals[vals > 5]                  # 7 11 9

# Replace outliers with NA
vals[vals > 10] = NA           # 4 7 NA 3 9
print(vals)

```
---

# Vectorized Iterations

R's `apply` family avoids explicit loops with concise, vectorized C-level execution.

```{r apply}

# Sample 4×3 matrix
mat = matrix(1:12, nrow = 4, byrow = TRUE)

# Column means
apply(mat, 2, mean)

# Row standard deviations
apply(mat, 1, sd)

```

```{r lapply}

set.seed(1942)

# 1. Create a list of score vectors -----------------------------
scores = lapply(1:5, \(i) rnorm(50, mean = 70 + i * 5, sd = 8))
names(scores) = paste0("Class_", 1:5)

# 2. Compute the mean score for every class ---------------------
class_means = lapply(scores, mean)

class_means

```

```{r mapply}

# Create 3 normally distributed vectors with different means and SDs
set.seed(1914)
params = data.frame(mu = c(0, 5, 10), sd = c(1, 2, 3))

samples = mapply(
  FUN = rnorm,
  n   = 5,                  # recycled for each call
  mean = params$mu,
  sd   = params$sd,
  SIMPLIFY = FALSE          # keep as list of vectors
)

str(samples)

```

```{r sapply}

nums = list(a = 1:4, b = 5:7, c = 8:9)

# Sums each element returning a numeric vector
sapply(nums, sum)

vec = 1:6
sapply(vec, function(x) x^2)

```

---

# Parallel Computing

```{r parallel-setup}
library(parallel)
n_cores = max(1L, detectCores() - 1L)
print(n_cores)
```

Use `parLapply` for cross-platform parallel execution, or `mclapply` for Unix-based systems.

```{r parallel-parlapply}

# Expensive function: Monte Carlo estimate of pi
estimate_pi = function(n = 1e6) {
  x = runif(n)
  y = runif(n)
  mean(x^2 + y^2 <= 1) * 4
}

n_rep = 30

# ---------------------------------------------------------
# 1. SEQUENTIAL RUN
# ---------------------------------------------------------
set.seed(1986)

t_seq = system.time({
  seq_results = lapply(1:n_rep, function(i) {
    t0 = proc.time()["elapsed"]
    pi_est = estimate_pi()
    
    # Return both the estimate and the time it took
    data.frame(run = i, pi_est = pi_est, elapsed = proc.time()["elapsed"] - t0)
  })
})

# Bind into a single data frame
seq_df = do.call(rbind, seq_results)

# ---------------------------------------------------------
# 2. PARALLEL RUN
# ---------------------------------------------------------
n_cores = max(1, detectCores() - 1)
cl = makeCluster(n_cores)
clusterExport(cl, varlist = "estimate_pi")

set.seed(1986)

t_par = system.time({
  par_results = parLapply(cl, 1:n_rep, function(i) {
    t0 = proc.time()["elapsed"]
    pi_est = estimate_pi()
    
    # Return both the estimate and the time it took
    data.frame(run = i, pi_est = pi_est, elapsed = proc.time()["elapsed"] - t0)
  })
})

stopCluster(cl)

# Bind into a single data frame
par_df = do.call(rbind, par_results)

# ---------------------------------------------------------
# 3. COMPARE RESULTS
# ---------------------------------------------------------
# Check if the estimates are mathematically equivalent
all.equal(seq_df$pi_est, par_df$pi_est)

# Print the timing results
cat("Total Sequential Time:     ", round(t_seq["elapsed"], 3), "seconds\n")
cat("Total Parallel Time:       ", round(t_par["elapsed"], 3), "seconds\n\n")

cat("Avg Sequential per run:    ", round(mean(seq_df$elapsed), 3), "seconds\n")
cat("Avg Parallel per run:      ", round(mean(par_df$elapsed), 3), "seconds\n")

```

```{r}

# Install if needed: install.packages(c("future", "future.apply"))
library(future)
library(future.apply)

# 1. Define the workhorse function
estimate_pi = function(n = 1e6) {
  mean(runif(n)^2 + runif(n)^2 <= 1) * 4
}

n_rep = 20

# 2. Tell R your "plan" (run asynchronously across multiple background sessions)
plan(multisession, workers = availableCores() - 1)

set.seed(1986)

# 3. Execute! future_lapply acts just like lapply, but routes traffic asynchronously
t_future = system.time({
  
  # future.seed = TRUE ensures random numbers are safely generated across cores
  future_results = future_lapply(1:n_rep, function(i) {
    t0 = proc.time()["elapsed"]
    pi_est = estimate_pi()
    data.frame(run = i, pi_est = pi_est, elapsed = proc.time()["elapsed"] - t0)
  }, future.seed = TRUE)
  
})

# 4. Bind the results
future_df = do.call(rbind, future_results)

# 5. Always explicitly close background workers when done
plan(sequential)

cat("Total Future Async Time: ", round(t_future["elapsed"], 3), "seconds\n")
cat("Average Future Async per run: ", round(mean(future_df$elapsed), 3), "seconds\n")

```
---

# Running Python Code in R

Execute Python scripts and pass data frames back and forth using `{reticulate}`.

```{r}

library(reticulate)

# Show which Python is being used
py_config()

#reticulate::py_install("pandas")

```


```{r reticulate-inline, eval=FALSE}

# Run Python inline
py_run_string("
x = [1, 2, 3]
y = [i**2 for i in x]
")

# Access Python variables natively in R
py$y
py$x

```
```{r}

np = import("numpy")

# Create NumPy arrays and compute statistics
a = np$array(c(1, 2, 3, 4, 5))
mean_a = np$mean(a)
mean_a

```

```{r}

# Create a standard R dataframe
r_df = data.frame(a = 1:3, b = c(4, 5, 6))

```

```{python}

import pandas as pd

# Pull the dataframe from R into Python
py_df = r.r_df

# Perform our Python operations
py_df['sum'] = py_df['a'] + py_df['b']

```

```{r}

# Pull the modified dataframe back from Python into R
final_r_df = py$py_df

print(final_r_df)

```
---

# GPU Computing in R

> *Note: The following code blocks are structured but intentionally unevaluated (`eval=FALSE`) in this notebook to prevent rendering failures on environments lacking CUDA/GPU configurations.*

> *Note on Windows Compatibility: Enabling GPU acceleration for TensorFlow in a Windows environment is not a "plug-and-play" process. This requires a specific infrastructure stack: a WSL2 (Windows Subsystem for Linux) installation, an Ubuntu R Server and the dedicated GPU version of TensorFlow. For a step-by-step walkthrough on configuring this environment, please refer to Chapter 6 of my Medium article, where I cover the full backend setup required to get R and your GPU talking to each other.*

## GPU Initialization

```{r gpu-verify, eval=FALSE}

library(tensorflow)
library(keras3)

# Check for the GPU
tf$config$list_physical_devices("GPU")

```

## Neural Network for Non-Linear Classification

```{r gpu-nn-config, eval=FALSE}

set.seed(1925)

n_per_class <- 2000

# Inner circle (class 0)
theta1 <- runif(n_per_class, 0, 2 * pi)
r1     <- rnorm(n_per_class, mean = 1, sd = 0.08)
x1     <- cbind(r1 * cos(theta1), r1 * sin(theta1))
y1     <- rep(0, n_per_class)

# Outer ring (class 1)
theta2 <- runif(n_per_class, 0, 2 * pi)
r2     <- rnorm(n_per_class, mean = 2, sd = 0.08)
x2     <- cbind(r2 * cos(theta2), r2 * sin(theta2))
y2     <- rep(1, n_per_class)

# Full dataset
x <- rbind(x1, x2)
y <- c(y1, y2)

# Shuffle
idx <- sample(seq_len(nrow(x)))
x   <- x[idx, ]
y   <- y[idx]

# Train / test split
n    <- nrow(x)
n_tr <- floor(0.8 * n)

x_train <- x[1:n_tr, ]
y_train <- y[1:n_tr]

x_test  <- x[(n_tr + 1):n, ]
y_test  <- y[(n_tr + 1):n]

model <- keras_model_sequential() %>%
  layer_dense(units = 32, activation = "relu", input_shape = 2) %>%
  layer_dense(units = 64, activation = "relu") %>%
  layer_dense(units = 32, activation = "relu") %>%
  layer_dense(units = 1, activation = "sigmoid")  # binary classification

model %>%
  compile(
    optimizer = "adam",
    loss      = "binary_crossentropy",
    metrics   = "accuracy"
  )

# summary(model)

batch_size <- 128
epochs     <- 20

t <- system.time({
  history <- model %>%
    fit(
      x = x_train, y = y_train,
      batch_size      = batch_size,
      epochs          = epochs,
      validation_split = 0.2,
      verbose         = 2
    )
})

cat("Training elapsed time (GPU-backed TF):", t["elapsed"], "seconds\n")

model %>%
  evaluate(x_test, y_test, verbose = 0)

```

```{r gpu-nn-plot, eval=FALSE}

library(ggplot2)
library(viridis)

# Define grid for visualization
x_min <- min(x[,1]) - 0.2
x_max <- max(x[,1]) + 0.2
y_min <- min(x[,2]) - 0.2
y_max <- max(x[,2]) + 0.2

grid <- expand.grid(
  x1 = seq(x_min, x_max, length.out = 400), # Higher resolution for crispness
  x2 = seq(y_min, y_max, length.out = 400)
)

# Predict probabilities
grid$prob <- as.numeric(model %>% predict(as.matrix(grid), verbose = 0))

ggplot() +
  # 1. Background Grid/Heatmap (Smooth Ukrainian Flag gradient)
  # Low prob (Inner) -> Blue. High prob (Outer) -> Yellow.
  geom_raster(data = grid, aes(x = x1, y = x2, fill = prob), interpolate = TRUE) +
  
  # 2. Add a strong decision boundary line
  geom_contour(data = grid, aes(x = x1, y = x2, z = prob), 
               breaks = 0.5, color = "white", linetype = "solid", linewidth = 1.2) +
  
  # 3. Add the points (Small and translucent so they don't hide the boundary)
  geom_point(aes(x = x[,1], y = x[,2], color = factor(y)), size = 0.6, alpha = 0.3) +
  
  # --- Color Palettes (Ukrainian Flag Theme) ---
  
  # Background Fill: Cividis provides a great smooth blue-to-yellow gradient
  scale_fill_viridis_c(option = "cividis", direction = -1,
                       name = "Predicted\nProbability", limits = c(0, 1),
                       breaks = c(0, 0.5, 1), labels = c("Inner", "Boundary", "Outer")) +
  
  # Points Color: Force them to pop against the blue/yellow background
  scale_color_manual(values = c("black", "red"), 
                     name = "True Class", labels = c("Inner (0)", "Outer (1)")) +
  
  # --- Theme and Labs ---
  labs(
    title = "Neural Network classification Boundary",
    subtitle = "Modern TensorFlow/Keras on GPU via WSL2",
    x = "Feature 1", 
    y = "Feature 2"
  ) +
  coord_fixed() + # Critical: keeps the circle circular
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    # Center the title and make it bold
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 12),
    plot.caption = element_text(size = 9, color = "gray50"),
    panel.grid.major = element_line(color = "gray90"),
    plot.background = element_rect(fill = "white", color = NA)
  )

)
```

---

# Session Info {.unnumbered}

```{r session-info}
sessionInfo()
```